Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-27 analysis/paper/paper.Rmd
fix typos and edit introduction section based on comments

The emergence of social inequality in the context of cross-cultural interaction, especially in a colonial situation, has been observed in many parts of world, for example, when foreign trade goods are introduced to local Indigenous societies. The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where European colonial rule was limited and their conquests were short-lived and unsuccessful, but their colonial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control and influence [@Trabert2017; @Acabado2017]. @Acabado2017's study of Ifugao society in the  Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation. Indigenous societies might not simply passively accept colonial rule, but instead actively negotiate with the colonist, and accommodate or resist foreign intrusion through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000]. In this paper we present a case study from Northeastern Taiwan as an example of these indirect impacts of foreign occupations on local Indigenous societies. Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies. Previous reports of archaeological evidence in northeastern Taiwan descrbes increases in the use of prestige goods, such as trade ornaments, and uneven artefact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century. Here we show that in addition to the differential distribution of trade ornaments post-European contact, the locally-made ceramics become highly consistent in their form and shape, compared to other pottery throughout northeastern Taiwan. This shift to higher consistency hints at the possibility of craft specialization in pottery manufacture.

To identify craft specialization in the archaeological record, variation in the standardization of a specialized product is often used as an indicator. This is based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, higher skills, and fewer producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For investigating ceramic standardization, several measurements have been discussed and used, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data. By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation of linear dimensions of pots, and differences between artisans can be detected at both intra-individual level and inter-individual level in a region. However, because pottery vessels are typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations. Thus, to capture subtle shape variations that might be relevant to standardization, a method that is senstive to non-linear aspects of shape variation is necessary. Here we analyse pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007]. Using geometric morphometry, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production resulting from cross-cultural interaction contexts.

Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000]. We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by icreased standardisation of pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality. In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by rgreater economic and social control by a small group of individuals at Kiwulan. Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia. Our findings improve our understanding of the relationship between the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf. @Acabado2017].

@@

However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks contains crucial variation. Furthermore, landmark points may be unavailable for structures that are mostly or entirely curves. In those cases, outline approaches are more suitable for analyzing the overall shape of an object. One outline approache is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010]. Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013]. Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].

EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation. Applications in archaeology include human remains and zooarchaeology to understand  biological variation, and more recently on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization. Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico. Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes. Among those highly standardized jars, some were made by attached specialists <!-- what does this mean, 'attached' to what? please clarify -->, while others were made by independent specialists according to their spatial distribution patterns. Similarity, @Selden2019 examined 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana <!-- using semi-landmarks or something else? please add -->. The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner. Also, two discrete base and body shapes represented northern and southern differences were identified. For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek201's EFA results help us understand the level of production standardization over time across the region. Those examples show that outline approaches can distinguish variation in ceramic shapes at high resolutions to understand ceramic production.

@@

Ceramics analyzed in this paper come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)). Kiwulan is situated on a hill near a riverside at the northern margin of Yilan Province, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides. Yilan is an ideal context to study peripheral colonial influence because Indigenous communities there were isolated by natural barriers that limited the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan. The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007]. This paper focuses on the upper component, dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period, which we define as the European presence in Taiwan in the early 17th century. The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 <!-- add citation for these dates -->. In 1642, the Spanish was expelled by the Dutch in northern Taiwan, who took over their forts at Helping Dau in Keelung, and in Tamsui. Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

@@

Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite (15-40%), metasandstone (1-10%),  sandstone fragments (1-6%), quartz (1-5%), and trace amounts of feldspar and slate. The size of the particles ranges from 500 micron to 1300 micron. In general, the globular vessel farbric presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz. The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016]. There are no substantial changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods.

@@

By standardizing standard deviation in each data set, the coefficients of variation values allows us to directly compare variation in samples measured with different units or means. We take this as our measurement of standardisation in vessel shape: lower CV values reflect higher standardisation, and thus increased craft specialisation in the community. To answer the question of whether CV values for vessels samples across our three occupational phases are significantly different or not, we used the the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014]. While previous work has used the @Feltz1996 asymptotic test for the equality of coefficients of variation from k populations <!-- add citation to archaeology papers using the asymptotic test  -->, we prefer the MSLR test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].

@@

To investigate the effect of the historically-documented decline of Indigenous population in the Chinese period <!-- add citation on population change during Chinese period --> on pottery production we examined the spatial distribution of pots in our sample in each period. Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialisation beause of reduced numbers of potters and pottery production shrinking a a few locations in the settlement during this time. Altough the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:pottery-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialsation.

>>>>

The emergence of social inequality in the context of cross-cultural interaction, especially in a colonial situation, has been observed in many parts of world, for example, when foreign trade goods are introduced to local Indigenous societies. The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where European colonial rule was limited and their conquests were short-lived and unsuccessful, but their colonial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control and influence [@Trabert2017; @Acabado2017]. @Acabado2017's study of Ifugao society in the  Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation. Indigenous societies might not simply passively accept colonial rule, but instead actively negotiate with the colonist, and accommodate or resist foreign intrusion through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000]. In this paper we present a case study from Northeastern Taiwan as an example of these indirect impacts of foreign occupations on local Indigenous societies. Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies. Previous reports of archaeological evidence in northeastern Taiwan describes increases in the use of prestige goods, such as trade ornaments, and uneven artefact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century. Here we show that in addition to the differential distribution of trade ornaments post-European contact, the locally-made ceramics become highly consistent in their form and shape, compared to other pottery throughout northeastern Taiwan. This shift to higher consistency hints at the possibility of craft specialization in pottery manufacture.

To identify craft specialization in the archaeological record, variation in the standardization of a specialized product is often used as an indicator. This is based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, higher skills, and fewer producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For investigating ceramic standardization, several measurements have been discussed and used, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data. By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation of linear dimensions of pots, and differences between artisans can be detected at both intra-individual level and inter-individual level in a region. However, because pottery vessels are typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations. Thus, to capture subtle shape variations that might be relevant to standardization, a method that is sensitive to non-linear aspects of shape variation is necessary. Here we analyse pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007]. Using geometric morphometry, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production resulting from cross-cultural interaction contexts.

Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000]. We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by increased standardisation of pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality. In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by greater economic and social control by a small group of individuals at Kiwulan. Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia. Our findings improve our understanding of the relationship between the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf. @Acabado2017].

@@

However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks contains crucial variation. Furthermore, landmark points may be unavailable for structures that are mostly or entirely curves. In those cases, outline approaches are more suitable for analyzing the overall shape of an object. One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010]. Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013]. Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].

EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation. Applications in archaeology include human remains and zooarchaeology to understand  biological variation, and more recently on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization. Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico. Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes. Among those highly standardized jars, some were made by attached specialists controlled by elites indicating the presence of social inequality, while others were made by independent specialists according to their spatial distribution patterns. Similarity, @Selden2019 examined 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana using semi-landmarks method. The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner. Also, two discrete base and body shapes represented northern and southern differences were identified. For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek201's EFA results help us understand the level of production standardization over time across the region. Those examples show that outline approaches can distinguish variation in ceramic shapes at high resolutions to understand ceramic production.

@@

Ceramics analyzed in this paper come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)). Kiwulan is situated on a hill near a riverside at the northern margin of Yilan Province, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides. Yilan is an ideal context to study peripheral colonial influence because Indigenous communities there were isolated by natural barriers that limited the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan. The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007]. This paper focuses on the upper component, dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period, which we define as the European presence in Taiwan in the early 17th century. The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 [@Andrade2007]. In 1642, the Spanish was expelled by the Dutch in northern Taiwan, who took over their forts at Helping Dau in Keelung, and in Tamsui. Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

@@

Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite (15-40%), metasandstone (1-10%),  sandstone fragments (1-6%), quartz (1-5%), and trace amounts of feldspar and slate. The size of the particles ranges from 500 micron to 1300 micron. In general, the globular vessel fabric presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz. The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016]. There are no substantial changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods.

@@

By standardizing standard deviation in each data set, the coefficients of variation values allows us to directly compare variation in samples measured with different units or means. We take this as our measurement of standardisation in vessel shape: lower CV values reflect higher standardisation, and thus increased craft specialisation in the community. To answer the question of whether CV values for vessels samples across our three occupational phases are significantly different or not, we used the the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014]. While previous work has used the @Feltz1996 Asymptotic test for the equality of coefficients of variation from k populations, we prefer the MSLR test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].

@@

To investigate the effect of the historically-documented decline of Indigenous population in the Chinese period [@Chen2007; @Hsieh2009] on pottery production we examined the spatial distribution of pots in our sample in each period. Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialisation because of reduced numbers of potters and pottery production shrinking a a few locations in the settlement during this time. Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:pottery-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialsation.
